---
title: "MapScape vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MapScape vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

MapScape is a visualization tool for spatial clonal evolution.

To run MapScape, type the following commands in R:

```{r, eval=FALSE}
install.packages("devtools") # if not already installed  
library(devtools)  
install_bitbucket("MO_BCCRC/mapscape")  
library(mapscape)  
example(mapscape) # to run examples
```

Three visualizations will appear in your browser (optimized for Chrome).

The first is of metastatic prostate cancer data published in Gundem et al. (2015):

```{r, echo=FALSE}
library(devtools)  
install_bitbucket("MO_BCCRC/mapscape")  
library(mapscape)  
# EXAMPLE 1 - Patient A21, Gundem et al., 2015

# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "A21_clonal_prev.csv", package = "mapscape"))
# mutations
mutations <- read.csv(system.file("extdata", "A21_mutations.csv", package = "mapscape"))
# locations of each tumour sample on user-provided image
sample_locations <- read.csv(system.file("extdata", "A21_sample_locations.csv", package = "mapscape"))
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "A21_tree.csv", package = "mapscape"))
# image reference
img_ref <- system.file("extdata", "A21_anatomical_image.png", package = "mapscape")
# radial order of samples
sample_ids <- c("H","F","J","D","A","I","C","E","G")
# run mapscape
mapscape(clonal_prev = clonal_prev, tree_edges = tree_edges, sample_locations = sample_locations, 
img_ref = img_ref, sample_ids = sample_ids, show_warnings=FALSE)
```

The second and third visualizations are of metastatic ovarian cancer patients 1 and 7 published in McPherson and Roth et al. (2015):

```{r, echo=FALSE}
# EXAMPLE 2 - Patient 1, McPherson and Roth et al., 2016
# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "px1_clonal_prev.csv", package = "mapscape"))
# mutations
mutations <- read.csv(system.file("extdata", "px1_mutations.csv", package = "mapscape"))
# locations of each tumour sample on user-provided image
sample_locations <- read.csv(system.file("extdata", "px1_sample_locations.csv", package = "mapscape"))
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "px1_tree.csv", package = "mapscape"))
# image reference
img_ref <- system.file("extdata", "px1_anatomical_image.png", package = "mapscape")
# colours for each clone
clone_colours <- data.frame( clone_id = c("A","B","C","D","E","F","G","H","I"), 
                             colour = c("d0ced0", "2CD0AB", "7FE9D1", "FFD94B", "FD8EE5", "F8766D", 
                             "4FD8FF", "B09AF5", "D4C7FC"))
# radial order of samples
sample_ids <- c("LFTB4", "LOvB2", "ApC1", "ROvA4", "ROv4", "ROv3", "ROv2", "ROv1", "RFTA16", "Om1", "SBwl", "SBwlE4")
# run mapscape
mapscape(clonal_prev = clonal_prev, tree_edges = tree_edges, sample_locations = sample_locations, 
img_ref = img_ref, clone_colours = clone_colours, sample_ids = sample_ids, show_warnings=FALSE)
```

```{r, echo=FALSE}
# EXAMPLE 3 - Patient 7, McPherson and Roth et al., 2016
# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "px7_clonal_prev.csv", package = "mapscape"))
# mutations
mutations <- read.csv(system.file("extdata", "px7_mutations.csv", package = "mapscape"))
# locations of each tumour sample on user-provided image
sample_locations <- read.csv(system.file("extdata", "px7_sample_locations.csv", package = "mapscape"))
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "px7_tree.csv", package = "mapscape"))
# image reference
img_ref <- system.file("extdata", "px7_anatomical_image.png", package = "mapscape")
# colours for each clone
clone_colours <- data.frame(clone_id = c("A","B","C","D","E"), 
                            colour = c("d0ced0", "2CD0AB", "FFD94B", "FD8EE5", "F8766D"))
# radial order of samples
sample_ids <- c("BwlA6", "RPvM", "RUtD1", "RUtD2", "RUtD3", "ROvC4", "ROvC5", "ROvC6", "LOv1","LOvA10","LOvA4","BrnM", "BrnMA1")
# run mapscape
mapscape(clonal_prev = clonal_prev, tree_edges = tree_edges, sample_locations = sample_locations, 
img_ref = img_ref, clone_colours = clone_colours, sample_ids = sample_ids, show_warnings=FALSE)
```
To view the documentation for MapScape, type the following command in R:

```{r, eval=FALSE}
?mapscape
```

MapScape was developed at the Shah Lab for Computational Cancer Biology at the BC Cancer Research Centre.
 
References:  

Gundem, Gunes, et al. "The evolutionary history of lethal metastatic prostate cancer." Nature 520.7547 (2015): 353-357.

McPherson, Andrew, et al. "Divergent modes of clonal spread and intraperitoneal mixing in high-grade serous ovarian cancer." Nature genetics (2016).